---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.path = "README-",
  message = FALSE, warning = FALSE, error = FALSE, tidy = TRUE)
```

# mapmate

[![Travis-CI Build Status](https://travis-ci.org/leonawicz/mapmate.svg?branch=master)](https://travis-ci.org/leonawicz/mapmate)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/leonawicz/mapmate?branch=master&svg=true)](https://ci.appveyor.com/project/leonawicz/mapmate)
[![Coverage Status](https://img.shields.io/codecov/c/github/leonawicz/mapmate/master.svg)](https://codecov.io/github/leonawicz/mapmate?branch=master)
[![Rdoc](http://www.rdocumentation.org/badges/version/mapmate)](http://www.rdocumentation.org/packages/mapmate)

`mapmate` (map animate) is an R package for map animation. It is used to generate and save a sequence of plots to disk as a still image sequence intended for later use in data animation production.

```{r is_md, echo = FALSE}
is_md <- knitr::opts_knit$get("rmarkdown.pandoc.to") == "markdown_github-ascii_identifiers"
```

```{r md_examples, echo = FALSE, results = "asis", eval = is_md}
cat("Here is the complete online [documentation and tutorials with code examples](https://leonawicz.github.io/mapmate/).")
```

## Installation and bug reporting

You can install snapverse from github with:

```{r gh-installation, eval = FALSE}
# install.packages('devtools')
devtools::install_github("leonawicz/mapmate")
```

Please file a minimal reproducible example of any clear bug at [github](https://github.com/leonawicz/mapmate/issues).

```{r html_examples, echo = FALSE, results = "asis", eval = !is_md}
cat(paste0(
  "## Example videos based on mapmate\n\n",
  "### Historical and projected global temperature anomalies\n\n",
  '<p align="center"><iframe src="https://www.youtube.com/embed/xhqEkyJDBho?ecver=2" width="640" height="360" frameborder="0" allowfullscreen></iframe></p>\n\n',
  "### Global UAF/SNAP Shiny Apps web traffic\n\n",
  '<p align="center"><iframe src="https://www.youtube.com/embed/uQYR91qixgo?ecver=2" width="640" height="360" frameborder="0" allowfullscreen></iframe></p>\n\n',
  "### Flat map great circle animation example\n\n",
  '<p align="center"><iframe src="https://www.youtube.com/embed/yoyIUMvIP3Q?ecver=2" width="640" height="360" frameborder="0" allowfullscreen></iframe></p>\n\n'
))
```

## Introduction and basic example

The `mapmate` package is used for map- and globe-based data animation pre-production.
Specifically, `mapmate` functions are used to generate and save to disk a series of map graphics that make up a still image sequence,
which can then be used in video editing and rendering software of the user's choice. This package does not make simple animations directly within R, which can be done with packages like `animation`.
`mapmate` is more specific to maps, hence the name, and particularly suited to 3D globe plots of the Earth.
Functionality and fine-grain user control of inputs and outputs are limited in the current package version.


```{r example_plots, eval=TRUE, fig.show='hold', fig.cap="2D flat map and 3D globe"}
library(mapmate)
library(dplyr)
data(annualtemps)
annualtemps

library(RColorBrewer)
pal <- rev(brewer.pal(11,"RdYlBu"))

temps <- mutate(annualtemps, frameID = Year - min(Year) + 1)
frame1 <- filter(temps, frameID==1) # subset to first frame
id <- "frameID"

save_map(frame1, z.name="z", id=id, ortho=FALSE, col=pal, type="maptiles", save.plot=FALSE, return.plot=TRUE)
save_map(frame1, z.name="z", id=id, col=pal, type="maptiles", save.plot=FALSE, return.plot=TRUE)

```

The above is only a very basic initial example of static 2D and 3D maps.
See the [introduction vignette](https://leonawicz.github.io/mapmate/articles/mapmate.html) for more complete and typical usage examples:

- Generate a data frame containing monthly map data (optionally seasonal or annual aggregate average data)
in the form of an n-year moving or rolling average based on an input data frame of raw monthly data.
- Generate a sequence of still frames of:
    - map data for use in a flat map animation.
    - dynamic/temporally changing map data projected onto a static globe (3D Earth)
    - static map data projected onto rotating globe
    - dynamic map data projected onto rotating globe
- Parallel processing examples using `mclapply`
- Convenient iterator wrapper function
- Comparison of map tiles, map lines, and polygons
- Non-map data example (time series line growth)

This and other vignettes covering more examples, including network maps with great circle arcs and generating video with `ffmpeg`,
can be found at the [mapmate](https://leonawicz.github.io/rvtable/articles/rvtable.html) website.

Other features and functionality will be added in future package versions.
